---
title: "EURO2016"
author: Tuomo Nieminen
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: cosmo
    df_print: paged
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "")
```

# Welcome

This GitHub page describes predicting the 2016 european soccer championships matches using ordinal regression. The GitHub repository related to this page is [here](https://github.com/TuomoNieminen/EURO2016) and an article in finnish describing the analysis is [here](http://tyyppiarvo.com/2016/06/jalkapallossa-kannattaa-laukoa-ennustimme-em-turnauksen-alkuvaiheen-ottelut-ja-jatkoonmenijat/)

# Qualifier data

## Features

```{r}

emq <- get(load("data/EUROq.Rda"))
emq <- emq[complete.cases(emq),]
names(emq)

```

## Correlations of ratios and differences

```{r}

# ratios and differences of team stats
homevars <- substr(names(emq),1,5)=="homeq"
awayvars <- substr(names(emq),1,5)=="awayq"
#fifa

homevars[7] <- T
awayvars[8] <- T
ratios <- emq[,homevars] / emq[,awayvars]
diffs <- emq[,homevars] - emq[,awayvars]

#correlations
corrs <- rbind(cor_ratios=apply(ratios,2,cor,emq$goal_diff),
               cor_diffs=apply(diffs,2,cor,emq$goal_diff))

t(corrs)
```

## Shot ratio versus goal difference

```{r}

# a scatter plot of shots ratio and goal difference
shot_ratio <- ratios[,4]
goal_diff <- emq$goal_diff
plot(shot_ratio, goal_diff,
     xlab="SR", 
     ylab="match goal difference",
     main="Qualifier shot ratio (SR) 
     versus tournament match goal difference 
     EM 2008 & 2012",
     col="green4", pch=20)
abline(lm(goal_diff~shot_ratio), col="darksalmon")
legend("bottomright",
       legend=paste("correlation:",round(cor(shot_ratio, goal_diff),2)),
       bty="n")

```


# Feature selection with regularized regression 

```{r, message = FALSE}

features <- cbind(ratios[,c(1,4)], diffs[,8])
colnames(features) <- c("fifa_ratio","shot_ratio","corners_diff")
outcome <- emq$outcome
library(glmnetcr)
netfit <- glmnet.cr(features, outcome, maxit=500)
```

## Bayesian information criteria

```{r}

par(mar=c(3,2,2,10), cex.axis=0.7)
plot(netfit, xvar="step", type="aic")
```

## Coefficient path

```{r}
par(mar=c(3,2,2,10), cex.axis=0.7)
plot(netfit, xvar = "step", type = "coefficients")
# only shot_ratio is meaningful
```

## Ordinal regression with shot ratio

```{r}
library(MASS)
fit1 <- polr(outcome ~ shot_ratio, Hess=T)
cbind.data.frame(outcome = outcome, 
                 prediction = predict(fit1, type="probs"))

```


# Uefa ratings

## Uefa rating ratio versus goal difference

```{r}

emu <- read.csv2("data/em2012_uefa.csv", stringsAsFactors = F)
emu$uefa_ratio <- emu$uefa_koti / emu$uefa_vieras
emu$goal_diff <- emu$maalit_koti - emu$maalit_vieras

# plot
uefa_ratio <- emu$uefa_ratio
goal_diff2 <- emu$goal_diff

plot(uefa_ratio,goal_diff2,
     xlab="UCR", 
     ylab="match goal difference",
     main="Uefa coefficient ratio (UCR) 
     versus tournament match goal difference \n  EM 2012.",
     col="green4", pch=20)
abline(lm(goal_diff2~uefa_ratio), col="darksalmon")
legend("bottomright",
       legend=paste("correlation:",round(cor(uefa_ratio, goal_diff2),2)),
       bty="n")

```

## Ordinal regression with uefa ratio

```{r}

# model based on 2012 uefa ratings
# define the outcome
emu$outcome <- 1
emu$outcome[emu$goal_diff > 0] <- 3
emu$outcome[emu$goal_diff < 0 ] <- 0
emu$outcome <- factor(emu$outcome,ordered = T, levels=c(0,1,3),labels=c("loss","draw","win"))

fit2 <- polr(outcome ~ uefa_ratio, Hess=T, data=emu)

cbind.data.frame(outcome = emu$outcome, 
                 prediction = predict(fit2, type="probs"))


```

# The 2016 matches and predictions

```{r}

# load 2016 matches
matches2016 <- get(load("data/matches2016.Rda"))

# use mass library to fit ordinal regression models
library(MASS)

# use the models to predict match outcome probabilities for 2016
probs1 <- predict(fit1, newdata=matches2016, type="probs")
probs2 <- predict(fit2, newdata=matches2016, type="probs")

# combine the predictions with the match data

predictions2016 <- 0.5*(probs1+probs2)[,3:1]
alkulohko <- cbind(matches2016, predictions2016)

# exclude the explanatory variables 
alkulohko2016 <- alkulohko[,c("date","group","hometeam","awayteam","win","draw","loss", "homeuefa", "awayuefa", "uefa_ratio", "homeavrg_shots", "awayavrg_shots", "shot_ratio")]
alkulohko2016
```

## Expected points per match

```{r}

# calculate expected points per match
getEV <- function(p) {sum(c(3,1,0)*p)}
homePoints <- apply(predictions2016, 1, getEV)
awayPoints <- apply(predictions2016[,3:1], 1, getEV)
alkulohko2016 <- cbind(alkulohko2016[, 1:7], homePoints, awayPoints)
# rounding
alkulohko2016[,5:9] <- round(alkulohko2016[,5:9], 2)

alkulohko2016
```

## Expected points by group

```{r}

# calculate expected points by group

#  function
get_total_points <- function(data, team) {
  homedata <- data[data$hometeam==team,]
  awaydata <- data[data$awayteam==team,]
  homeEV <- sum(homedata$homePoints)
  awayEV <- sum(awaydata$awayPoints)
  return(homeEV+awayEV)
}

group <- alkulohko2016$group

#  group predictions
by(alkulohko2016,group,FUN=function(d) {
  teams <- unique(c(d$hometeam,d$awayteam))
  points <- t(sort(sapply(teams,get_total_points,data=d),decreasing = T))
  rownames(points) <- "points"
  points
})

```

<br>

<hr>

Tuomo Nieminen 2017

<br>